---
title: "HW3_MarkMalysa"
output: html_document
date: "2023-10-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r importingCSV}
USlifehistory <- read.csv('USlifehistory.csv')
galton_height <- read.csv('galton_height.csv')
galton_peas <- read.csv('galton_peas.csv')
```

```{r librarys}
library(skimr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(moderndive)
```

```{r 1-1 QQ plot for years since 1880}
yearsfrom1880 <- USlifehistory$year-1880
score_model1 <-lm(life_expectancy~yearsfrom1880, data= USlifehistory)
score_model1
summary(score_model1)
get_regression_table(score_model1)
residuals <- score_model1$residuals
qq <- ggplot(data = USlifehistory, mapping = aes(sample = residuals)) +
  stat_qq() +
  labs(title = "QQ Plot of Residuals",
       x = "Quantiles",
       y = "Residuals")
qq
```


```{r 1-2 histogram of residual and scatter plot of residual vs. x}
hist(residuals)
ggplot(data=USlifehistory, aes(x=yearsfrom1880, y=residuals)) + geom_point() + geom_smooth(method = lm)
```
From the scatter plot and the mean of the residuals one can see that their is a linear pattern between the y and the x. This can be seen from the blue line being approximately horizontal at y = 0. Looking at the scatter plot because there seems to be some sort of a systemic pattern once can not assume independence. Moving on, from the shape of the histogram one can say that this is approximately normal and therefore one can assume normality. Looking at the overall distribution of the scatter plot one can say that there is homoscedasticity because of the equal variance. 

```{r 2 Regression of child’s height (gender adjusted) on mid-height of parent}
galton_height$mid_Parent_height <- (galton_height$Father + galton_height$Mother)/2

meanratio <- mean(galton_height$Height[galton_height$Gender == "M"]) / mean(galton_height$Height[galton_height$Gender == "F"])


galton_height <- galton_height %>%
  mutate(
    adj_height_Mother = Mother*meanratio
  )

galton_height <- galton_height %>%
  mutate(
    adj_height = ifelse(Gender == "F", (Height*meanratio), Height)
  )
galton_height$mid_Parent_height <- (galton_height$Father + galton_height$adj_height_Mother)/2


```

```{r 2-1 Have a scatterplot of y vs. x}
ggplot(data=galton_height, aes(x=mid_Parent_height, y=adj_height)) + geom_point() + geom_smooth(method = lm) + geom_abline(intercept = 0, slope = 1, color = "red") 
```
```{r 2-2 What is the average children’s height in the data}
mean(galton_height$adj_height)
mean(galton_height$mid_Parent_height)
```


```{r 2-3 Among parents whose mid-height between 72 and 73 inches...}
mean(galton_height$adj_height[galton_height$mid_Parent_height < 73 & galton_height$mid_Parent_height > 72])

```

```{r 2-4 Run regression, is the model significant?}
regressModel1 <- lm(data = galton_height, adj_height~mid_Parent_height)
summary(regressModel1)
plot(regressModel1)
```
As seen in the regression model the p-value is less than 0.05, which means this model is significant. 

#2-5 If the parents’ mid-height increases by 1 inch ..
As seen from the summary of the regression model an inch increase in the parents mid height will result in an expected 0.72894 inch increase in the childs adjusted height. This expected increase is less than 1 inch. 

```{r 2-6 Estimate the child’s height if the mid-height of parent is 64, 68, 70, 72, 76...}
regressModel1 <- lm(data = galton_height, adj_height~mid_Parent_height)
newValues <- data.frame(mid_Parent_height = c(64, 68, 70, 72, 76))
new_estimates <- predict(regressModel1, newValues)
new_estimates

new_estimates - mean(galton_height$adj_height)
(new_estimates - mean(galton_height$adj_height))/sd(galton_height$adj_height)
```
In the above result of the code, 1 is paired with 64, 2  is paired with 68, 3 is paired with 70, 4 is paired with 72, and 5 is paired with 76. The row underneath that represents the difference between the predicted value and the mean. The last row represents how many standard deviations the predicted values are from the mean value. 


```{r 3 Regression of mid-height of parent on child’s height (gender adjusted)}
regressModel2 <- lm(data = galton_height, mid_Parent_height~adj_height)
```

```{r 3-1 Have a scatterplot of y vs. x}
ggplot(data=galton_height, aes(x=adj_height, y=mid_Parent_height)) + geom_point() + geom_smooth(method = lm) + geom_abline(intercept = 0, slope = 1, color = "red")
```
```{r 3-2 Among all children with height between 72 and 73 inches...}
mean(galton_height$mid_Parent_height[galton_height$adj_height < 73 & galton_height$adj_height > 72])
```

```{r 3-3 Run regression, is the model significant?}
regressModel2 <- lm(data = galton_height, mid_Parent_height~adj_height)
summary(regressModel2)
plot(regressModel2)
```
As seen in the above summary the p-value is less than 0.05 making the model significant. 

#3-4 If the child’s height increases by 1 inch..
If the child’s height increases by 1 inch the parents mid height will increase by 0.35744 inches which is smaller than 1 inch. 

```{r 3-5  Estimate the child’s height if the mid-height of parent is 64, 68, 70, 72, 76...}

newValues2 <- data.frame(adj_height = c(64, 68, 70, 72, 76))
new_estimates2 <- predict(regressModel2, newValues2)
new_estimates2

new_estimates2 - mean(galton_height$mid_Parent_height)
(new_estimates2 - mean(galton_height$mid_Parent_height))/sd(galton_height$mid_Parent_height)
```
In the above result of the code, 1 is paired with 64, 2  is paired with 68, 3 is paired with 70, 4 is paired with 72, and 5 is paired with 76. The row underneath that represents the difference between the predicted value and the mean. The last row represents how many standard deviations the predicted values are from the mean value. 


# Use the above results to explain regression to the mean.
So the above results essentially prove the swap between extreme cases as show by the two regression models. If a child's adjusted average height is more of an extreme case, the mid height of the parents is more likly to be closer to the mean. This is the same for the mid height of the parents, since if it is more extreme than the child's adjusted average height is more likly to be average. This can be seen in results of the two models and the visaulisations of the 2 models. 
